---
title: "Analyze ham radio callsign suffixes"
author: "Andrew Zimolzak"
date: '2022-05-11'
output: pdf_document
---

```{r libraries, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(data.table)
library(tidyr)
library(knitr)  # kable
```


# Input data and show size

```{r reading}
# message("Reading dat (~8 sec...)")
d_1528k = fread(here('EN.dat'), header = FALSE, sep="|", quote="")
d_100k = fread(here('ENshort.dat'), header = FALSE, sep="|", quote="")
```


```{r manual-set-data-size}
X = d_1528k
dim(X)
```


```{r rename}
X %>%
  rename(
    tablename = V1, num = V2, callsign = V5, code = V6, lnum = V7, fullname = V8,
    firstname = V9, middlename = V10, lastname = V11, suffix = V12, address = V16,
    city = V17, state = V18, zip = V19, weird_addr_text = V20, misctext = V21,
    zeronum = V22, othernum = V23, othercode = V24) -> renamed
```


# String processing of suffixes

```{r suffix-wait}
renamed %>%
  select(callsign) %>%
  separate(
    callsign, sep='[0-9]', into=c('prefix', 'suffix'), extra='merge'
  ) %>%
  mutate(len_suf = nchar(suffix)) %>%
  filter(len_suf >= 3) %>%
  mutate(
    s1 = substr(suffix, 1, 1),
    s2 = substr(suffix, 2, 2),
    s3 = substr(suffix, 3, 3)
  ) -> presuf

dim(presuf)
dim(presuf)[1] / dim(renamed)[1]
dim(presuf)[1] / (26 ^ 3)
```

The above shows: How many hams have a 3-char suffix, what proportion of hams
have a 3-char suffix, and how many hams per suffix (mean).

```{r onevar}
hams_per_suff <- dim(presuf)[1] / (26 ^ 3)
```



# Count unique suffixes

```{r distribution}
presuf %>%
  arrange(suffix) %>%
  group_by(suffix, s1, s2, s3) %>%
  summarise(count = n()) %>%
  ungroup() -> counts

26 ^ 3 - dim(counts)[1]
```

The above shows how many suffixes are *missing,* or not attested, out of the
expected $26^3 = 17576$.


## Most common suffixes

```{r print-kable-top}
counts %>%
  arrange(desc(count), suffix) %>%
  select(suffix, count) %>%
  head(n=35) %>%
  kable()
```


# Uncommon suffixes

```{r expand}
counts %>%
  filter(s3 != '0') %>%
  expand(s1, s2, s3) -> expanded

left_join(expanded, counts, by=c('s1', 's2', 's3')) -> counts_all

dim(expanded)
dim(counts_all)
26^3
```

Compare dimension of `expanded` and `counts_all` with the expected $26^3$.
Expect to see 17,576 show up in all of those. **Funny thing:** Now I am
seeing 18,252 in the `expanded` data frame. Why? That equals
$26 \cdot 26 \cdot 27$. Must be extra weird character in `s1` or `s2` or
`s3`. Added `filter() %>%` so now probably **fixed.**


## Table of never-used suffixes

```{r filter-na}
counts_all %>%
  filter(is.na(count)) -> forbidden

dim(forbidden)[1]
26 ^ 3 - dim(counts)[1]
```

Expect the above two to be equal. But again, weird. I am seeing 765
versus 89. It so happens that $765 - 89 = 26^3$. Seems like `s3` is
sometimes equal to the string "0".

**Final weird thing:** Why am I seeing 90 in `forbidden` versus 89 missing
from `counts`, even after filtering out the zero versus letter O thing?

## Diagnostic

```{r diagnostic}
counts %>% filter(s3 == 0) %>% kable()
```

I even wonder if this is a typo in the database of "zero" replacing
"capital letter O, as in Oscar."




# Final tables

Lots of suffixes starting in "Q" are banned. But *only three others*
are banned.

```{r non-q}
forbidden %>% filter(s1 != 'Q') %>% kable()
```

So, somewhat surprisingly to me, all of the following exist at least once:
ASS, POO, FUK.

```{r sketchy-suffixes}
counts %>%
  filter(
    suffix == 'ASS' | suffix == 'POO' | suffix == 'KKK' | suffix == 'FUK' |
      suffix == 'TIT' | suffix == 'GOD' | suffix == 'FCC' | suffix == 'XXX' |
      suffix == 'SEX' | suffix == 'PEE' | suffix == 'KOK' | suffix == 'NSA' |
      suffix == 'DIK' | suffix == 'DIC' | suffix == 'CIA' | suffix == 'ZZZ' |
      suffix == 'ZYX' | suffix == 'USA' | suffix == 'WTF'
  ) %>%
  arrange(desc(count)) %>%
  mutate(obs_over_exp = count / hams_per_suff) %>%
  kable()
```

"POO" is one of the few mildly rude ones that is actually *more* popular than
expected. Yikes at the substring with 68 callsigns containing it.


```{r q-with-rstu}
forbidden %>%
  filter(s1 == 'Q') %>%
  group_by(s1, s2) %>%
  summarise(sum_q = n()) %>%
  kable()
```

You can't have any suffix starting with "QR," "QS," or "QT." Plus a few starting with "QU."

```{r qu}
forbidden %>%
  filter(s1 == 'Q' & s2 == 'U') %>%
  kable()
```

## Diagnostic

```{r count-89-90}

forbidden %>% mutate(suf_join = paste(s1, s2, s3, sep='')) -> forbidden_joinable

left_join(
  forbidden_joinable,
  counts,
  by = c('suf_join' = 'suffix'),
  suffix = c('_for', '_cou')
) %>%
  select(suf_join, suffix, count_for, count_cou)
```

```{r math}
forb_n <- 3  + 26*3 + 9
forb_n
26^3 - forb_n
dim(counts)
```
